Visualizing Contingency Tables

### 2006-09-30 trunk/vcd/paper/Handbook/hsc.Rnw
final edits from Antony Unwin
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/vcd@578 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

\noindent Current implementations of mosaic displays can be found,
e.g., for \proglang{SAS} \cite{vcd:SAS:2005}, \pkg{ViSta} \cite{vcd:young:1996}, 
\pkg{MANET} \cite{vcd:Unwin+Hawkins+Hofmann:1996},
\pkg{Mondrian} \cite{vcd:Theus:2003},
\proglang{R} \cite{vcd:R:2005}, and \proglang{S-PLUS} \cite{vcd:SPLUS:2005}. 
Implementations of association and sieve plots can only be found in
\proglang{R} and \proglang{SAS} (in the latter, these plots are available
for two-way tables only).
Table~\ref{tab:compare} gives an overview of the available functionality in these systems.
The figures in this chapter have all been produced using the
\proglang{R} \index{R} system, using the extension packages \pkg{vcd} 
\cite{vcd:Meyer+Zeileis+Hornik:2005} and
\pkg{scatterplot3d} \cite{vcd:ligges+maechler:2003}
(Figure~\ref{fig:bars3d} only), all freely available from the
Comprehensive \proglang{R} Archive Network
(\url{http://CRAN.R-project.org/}). The \proglang{R} code used for the
figures is available from \url{http://statmath.wu-wien.ac.at/projects/vcd/}.

@@

Since mosaic plots are asymmetric by construction, the choice of the
variable order matters, as the first splitting variable dominates the 
plot. In our example, if we use `Visit frequency' %'
as the first splitting variable, the impression is very different compared to the previous
mosaic (see Figure~\ref{fig:mosaic3fig}). 
In this alternative display, we see the marginal distribution of `Visit frequency' %'
in the rows: about half of the patients get visited regularly. This
group is dominated by patients staying between 2 and 9 years.
It seems apparent that the distribution of LOS %'
is similar for monthly and never visited patients, so this two categories actually
represent one homogenous group (patients visited only casually). Since
the first splitting variable dominates the plot, it should be chosen
as to be the explanatory variable.

@@

\noindent It is easy to compute the \emph{expected} table under either
of these hypotheses. To fix notations, in the following we consider a 
2-way contingency table with $I$ rows and $J$ columns,
cell frequencies $\{n_{ij}\}$ for
$i = 1, \dots, I$ and $j = 1, \dots, J$, and row and column sums
$n_{i+} = \sum_i n_{ij}$ and $n_{+j} = \sum_j n_{ij}$, respectively. For
convenience, the number of observations is denoted $n = n_{++}$.
Given an underlying distribution with theoretical cell probabilities
$\pi_{ij}$, the null hypothesis of independence of the two categorical
variables can be formulated as
\begin{equation} \label{eq:H0}
H_0: \; \pi_{ij} \quad = \quad \pi_{i+} \pi_{+j}.
\end{equation}

@@

Now, the expected cell frequencies in this model are simply
$\hat n_{ij} = n_{i+}n_{+j}/n$.
The expected table for our sample data is given in Table~\ref{tab:hospital2}.
It could again be visualized using a mosaic plot, this
time applied to the table of expected frequencies. 
If we cross-tabulate each tile to fill it with a number of squares equal to
the corresponding number of \emph{observed} frequencies, we get a \emph{sieve plot}
(see Figure~\ref{fig:sieve}). It implicitly compares
expected and observed values since the density of the grid will
increase with the deviation of the observed from the expected
values. This allows the detection of general association patterns (for
nominal variables) and of linear association (for ordinal variables),
the latter producing tiles of either very high or very low density 
along one of the diagonals.
In the case of our data, the density of the rectangles is marked along the secondary
diagonal, indicating a negative association of 
the two variables. This gives evidence to the fact that for these patients,
visit frequency decreases with the length of stay.

@@

\noindent which are standardized raw residuals. In an 
association plot \cite{vcd:Cohen:1980}, each cell is represented 
by a rectangle that has (signed) height
proportional to the corresponding Pearson residual $r_{ij}$ and width proportional to the
square root of the expected counts $\sqrt{\hat n_{ij}}$. Thus, the area is proportional to
the raw residuals $n_{ij} - \hat n_{ij}$. The sign is visualized by
its position relative to the baseline (upward tiles for positive, and downward 
tiles for negative residuals). Figure~\ref{fig:assoc}
shows the association plot for the hospital data. 
Consistent with the corresponding mosaic and sieve plots, 
we clearly see that too many (too few) patients that stay from 
2 to 9 (more than 19) years get visited regularly
than would be expected under the null of independence, and that this pattern is 
reversed for patients visited less than monthly and that get never visited.

@@

As introduced in the previous section for association plots, the
investigation of residuals from a posited independence model is of
major interest in analyzing contingency tables. In the following, we
will demonstrate how the use of colors can greatly facilitate the
detection of interesting patterns. Before, we start with some general 
remarks on colors and color palettes.

@@

Choosing colors and color palettes in RGB space can be a bit inconvenient
and hence software implementations are often based on \index{color space!HSV}
Hue-Saturation-Value (HSV) colors (or the comparable Hue-Luminance-Satu\-ration 
color scheme). Both spaces are rather similar transformations of the RGB space
\cite{vcd:Brewer:1999,vcd:Poynton:2000} and are very common implementations of
colors in many computer packages \cite{vcd:Moretti+Lyons:2002}. Each color in HSV space is
represented by three dimensions $(H, S, V)$: the hue $H$ (dominant wavelength in the spectrum, in $[0, 360]$), 
the saturation $S$ (`colorfulness' or `pureness', in $[0, 100]$), and 
the value $V$ (`brightness', amount of gray, in $[0, 100]$).\footnote{In many implementations,
all three dimensions are scaled to the unit interval instead of the coordinates used here.}
These intuitive dimensions
make HSV colors easyier to specifiy than, e.g, RGB colors. 
However, HSV colors have several disadvantages.
Most importantly, HSV colors are not perceptually
uniform because the three HSV dimensions map only poorly to the three
perceptual dimensions of the human visual system  \cite{vcd:Brewer:1999,vcd:Ihaka:2003}.
One important issue here is that HSV dimensions are confounded, e.g., saturation
is not uniform across different hues.
As an example, see Figure~\ref{fig:colors} (left) 
showing a qualitative color palette \index{color palette!qualitative}
(colors $(H, 100, 100)$ for varying hues $H$) in the HSV
space: although saturation and value are fixed, the fully saturated blue is
perceived much darker than the fully saturated red or green,
making it difficult to judge the size of shaded areas.
Furthermore, the flashy fully saturated HSV colors are hard 
to look at for a longer time. For similar reasons, it is equally difficult
to derive acceptable diverging palettes from the HSV space, i.e.,
\index{color palette!diverging}
bipolar scales containing colors ranging between two very distinct colors.
The upper part of Figure~\ref{fig:diverging} shows a diverging
palette in the HSV space with colors ranging from a saturated red $(0, 100, 100)$
over a neutral white $(H, 0, 100)$ to a saturated blue $(240, 100, 100)$.
Although the palette should be balanced with
respect to colorfulness and brightness, the red colors are perceived
to be more more intense and flashy than the corresponding blue
colors.

@@

\setkeys{Gin}{width=0.7\textwidth}
\begin{figure}
\begin{center}
<<highlight,fig=TRUE,echo=FALSE,height=6,width=7,results=hide>>=
mosaic(t(hospital), 
       mar = c(left = 3.5), 
       labeling_args = list(offset_labels = c(left = 0.5), 
         offset_varnames = c(left = 1, top = 0.5), set_labels = 
         list("Visit frequency" = c("Regular","Less than\nmonthly","Never"))),
       split = TRUE, spacing = spacing_highlighting, 
       gp = gpar(fill = rep(grey.colors(2)[2:1], 1:2)))
@ 
\caption{Spine plot with highlighting for the hospital data.}
\label{fig:highlight}
\end{center}
\end{figure}

@@

A first step in the exploratory analysis of more complex tables is to
get a quick overview on the data. For this, 
all basic plots can be combined in pairwise displays,
arranged in a matrix similar to scatterplots in a pairs
plot. \index{pairs plot}
The diagonal cells contain the variable names, optionally with
univariate statistics, whereas the off-diagonal cells feature plots whose
variables are implicitly specified by the cells' position in the matrix. %'
In Figure~\ref{fig:pairs}, the diagonal cells show bar plots for the
distribution of the variables, and the off-diagonal cells mosaic plots
for the corresponding pair of variables. The plots suggest that
admission differs between male and female students, and between the departments,
and that the proportion of male and female students varies across the departments.
In particular, departments A and B have a higher proportion of male students
and a lower rejection rate compared to the other departments.

@@

The next step is to investigate three-way interactions among the variables.
In this example, we have a binary variable of interest (admission)
that shall be ``explained'' by the others. 
A natural way of representing such a three-way table is to use a mosaic
display, 
splitting first in the explanatory variables department and gender and highlighting
the resulting mosaic with respect to the dependent variable admissions. Here, we
use the vertical splits for both explanatory variables, resulting in the 
\index{doubledecker plot} doubledecker plot 
in Figure~\ref{fig:doubleucb}. From the widths of the
tiles it can be seen that students apply to the six departments in
unequal numbers (with a particularly small
number of females in departments~A and B),
and from the highlighting we see again that the admission rates differ
among the departments (roughly speaking, the admission rate is high for A and B,
low for F and in between for C to E). The rates 
are equal for male and female students, except for department A where
\textit{more} female than male students are admitted.

@@

In a first step, we create a mosaic plot for an exploratory view on the data:
Attitude towards punishment clearly is the dependent variable 
here and should be used as the last split via highlighting. 
Furthermore, we expect age and education to have an influence on both the memory
and the attitude, hence they should be used first for splitting as stratifying
variables. In fact, the question is whether memory has some influence on attitude, given 
age and education. We choose to create a three-way mosaic plot of the 
hypothesized explanatory variables (first splitting horizontically by age, 
then vertically by education, and finally horizontically again by memory), 
and to have attitude, the dependent variable, 
highlighted in the tiles (see Figure~\ref{fig:punishhigh}).
We can see that half of the people have more than 40 years of
age, most of which of only completed elementary school (much more than 
in the other age groups). Furthermore, it can be seen that the proportion
of people with memories of punishments increases with age and that
the approval rate decreases with education (this, however, is better shown in 
Figure~\ref{fig:punishcotab} described below). As for the question whether
attitude depends on memory, the plot suggests quite clearly (and somewhat
surprisingly) that for those people with elementary school education,
a higher proportion \emph{with} memories tends to
accept moderate punishment of children than those without: 
experienced violence seems to engender violence again.
For the other education groups, the picture is less clear: some cells 
indicate the same association while others do not.

@@

In summary, both age and education clearly matter here.
But is there really any significant influence of memory on the attitude towards 
punishment apart from that? Or, rephrased as a hypothesis:
Are memory and attitude conditionally independent, given age and education?
To focus on this question, we take another model-based view on the data
which avoids the distracting influence of the marginal variables (age
and education): We employ
partial mosaic plots in a trellis layout (conditioning on age and education) with
residual-based shading and data-driven cut-offs (see Figure~\ref{fig:punishcotab}).
This visualization brings out much more clearly that the increased approval rate
among people with memories of punishments is present mainly in the elementary education
column and, to a lesser degree, in the age 40+ row. Using, e.g., a 
permutation test similar to the one discussed in Section \ref{sec:test}, 
model statistics reveal that the hypothesis is rejected at a significance level of 
1\%, but the shading clearly shows that this association is
only significant in two cells (the two older age groups with elementary education)
where fewer people without memories do not approve punishment than expected under
(conditional) independence. In the group of people with age between
25--39, the lighter shading indicates that here,
the association is only significant at the 10\% level.

@@

The advantage of the exploratory view 
is to visualize the joint distribution of all variables.
On the other hand, the visualization might be 
strongly influenced by the marginal distribution of education over age (in particular
the large proportion of 40+ people with elementary education)
which is not of interest in the conditional independence problem. The
partial mosaic plot suppresses this effect by complementing the 
conditioning in the model with conditioning in the visualization.

@@

Mosaic, association, and sieve plots can be used for visualizing multi-way tables
by applying them to flat representations. Mosaic plots are particularly useful
for exploratory analysis whereas the other two require specification of a certain
independence model for which deviations should be brought out. In addition, specialized
`flavours' of the plots can leverage the
exploratory (pairs plots, highlighting, doubledecker plots) 
or model-based analysis (residual-based shading, conditional plots).

>>>>

\noindent Current implementations of mosaic displays can be found,
e.g., for \proglang{SAS} \cite{vcd:SAS:2005}, \pkg{ViSta} \cite{vcd:young:1996}, 
\pkg{MANET} \cite{vcd:Unwin+Hawkins+Hofmann:1996},
\pkg{Mondrian} \cite{vcd:Theus:2003},
\proglang{R} \cite{vcd:R:2006}, and \proglang{S-PLUS} \cite{vcd:SPLUS:2005}. 
Implementations of association and sieve plots can only be found in
\proglang{R} and \proglang{SAS} (in the latter, these plots are available
for two-way tables only).
Table~\ref{tab:compare} gives an overview of the available functionality in these systems.
The figures in this chapter have all been produced using the
\proglang{R} \index{R} system, using the extension packages \pkg{vcd} 
\cite{vcd:Meyer+Zeileis+Hornik:2006} and
\pkg{scatterplot3d} \cite{vcd:ligges+maechler:2003}
(Figure~\ref{fig:bars3d} only), all freely available from the
Comprehensive \proglang{R} Archive Network
(\url{http://CRAN.R-project.org/}). The \proglang{R} code used for the
figures is available from \url{http://statmath.wu-wien.ac.at/projects/vcd/}.

@@

Since mosaic plots are asymmetric by construction, the choice of the
variable order matters, as the first splitting variable dominates the 
plot. In our example, if we use `Visit frequency' %'
as the first splitting variable, the impression is very different compared to the previous
mosaic (see Figure~\ref{fig:mosaic3fig}). 
In this alternative display, we see the marginal distribution of `Visit frequency' %'
in the rows: about half of the patients get visited regularly. This
group is dominated by patients staying between 2 and 9 years.
It seems apparent that the distribution of LOS %'
is similar for monthly and never visited patients, so this two categories actually
represent one homogenous group (patients visited only casually). Since
the first splitting variable dominates the plot, it should be chosen
to be the explanatory variable.

@@

\noindent It is easy to compute the \emph{expected} table under either
of these hypotheses. To fix notations, in the following we consider a 
2-way contingency table with $I$ rows and $J$ columns,
cell frequencies $\{n_{ij}\}$ for
$i = 1, \dots, I$ and $j = 1, \dots, J$, and row and column sums
$n_{i+} = \sum_j n_{ij}$ and $n_{+j} = \sum_i n_{ij}$, respectively. For
convenience, the number of observations is denoted $n = n_{++}$.
Given an underlying distribution with theoretical cell probabilities
$\pi_{ij}$, the null hypothesis of independence of the two categorical
variables can be formulated as
\begin{equation} \label{eq:H0}
H_0: \; \pi_{ij} \quad = \quad \pi_{i+} \pi_{+j}.
\end{equation}

@@

Now, the expected cell frequencies in this model are simply
$\hat n_{ij} = n_{i+}n_{+j}/n$.
The expected table for our sample data is given in Table~\ref{tab:hospital2}.
It could again be visualized using a mosaic plot, this
time applied to the table of expected frequencies. 
If we cross-tabulate each tile to fill it with a number of squares equal to
the corresponding number of \emph{observed} frequencies, we get a \emph{sieve plot}
(see Figure~\ref{fig:sieve}). It implicitly compares
expected and observed values since the density of the grid will
increase with the deviation of the observed from the expected
values. This allows the detection of general association patterns (for
nominal variables) and of linear association (for ordinal variables),
the latter producing tiles of either very high or very low density 
along one of the diagonals.
In the case of our data, the density of the rectangles is marked along the secondary
diagonal, indicating a negative association of 
the two variables. This gives evidence for the fact that for these patients,
visit frequency decreases with the length of stay.

@@

\noindent which are standardized raw residuals. In an 
association plot \cite{vcd:Cohen:1980}, each cell is represented 
by a rectangle that has (signed) height
proportional to the corresponding Pearson residual $r_{ij}$ and width proportional to the
square root of the expected counts $\sqrt{\hat n_{ij}}$. Thus, the area is proportional to
the raw residuals $n_{ij} - \hat n_{ij}$. The sign is visualized by
its position relative to the baseline (upward tiles for positive, and downward 
tiles for negative residuals). Figure~\ref{fig:assoc}
shows the association plot for the hospital data. 
Consistent with the corresponding mosaic and sieve plots, 
we clearly see that too many (too few) patients that stay from 
2 to 9 (more than 19) years get visited regularly
than would be expected under the null of independence, and that this pattern is 
reversed for patients visited less than monthly or never visited.

@@

As introduced in the previous section for association plots, the
investigation of residuals from a posited independence model is of
major interest in analyzing contingency tables. In the following, we
will demonstrate how the use of colors can greatly facilitate the
detection of interesting patterns. We start with some general 
remarks on colors and color palettes.

@@

Choosing colors and color palettes in RGB space can be a bit inconvenient
and hence software implementations are often based on \index{color space!HSV}
Hue-Saturation-Value (HSV) colors (or the comparable Hue-Luminance-Satu\-ration 
color scheme). Both spaces are rather similar transformations of the RGB space
\cite{vcd:Brewer:1999,vcd:Poynton:2000} and are very common implementations of
colors in many computer packages \cite{vcd:Moretti+Lyons:2002}. Each color in HSV space is
represented by three dimensions $(H, S, V)$: the hue $H$ (dominant wavelength in the spectrum, in $[0, 360]$), 
the saturation $S$ (`colorfulness' or `pureness', in $[0, 100]$), and 
the value $V$ (`brightness', amount of gray, in $[0, 100]$).\footnote{In many implementations,
all three dimensions are scaled to the unit interval instead of the coordinates used here.}
These intuitive dimensions
make HSV colors easier to specify than RGB colors. 
However, HSV colors have several disadvantages.
Most importantly, HSV colors are not perceptually
uniform because the three HSV dimensions map only poorly to the three
perceptual dimensions of the human visual system  \cite{vcd:Brewer:1999,vcd:Ihaka:2003}.
One important issue here is that HSV dimensions are confounded, e.g., saturation
is not uniform across different hues.
As an example, see Figure~\ref{fig:colors} (left) 
showing a qualitative color palette \index{color palette!qualitative}
(colors $(H, 100, 100)$ for varying hues $H$) in the HSV
space: although saturation and value are fixed, the fully saturated blue is
perceived much darker than the fully saturated red or green,
making it difficult to judge the size of shaded areas.
Furthermore, the flashy fully saturated HSV colors are hard 
to look at for a longer time. For similar reasons, it is equally difficult
to derive acceptable diverging palettes from the HSV space, i.e.,
\index{color palette!diverging}
bipolar scales containing colors ranging between two very distinct colors.
The upper part of Figure~\ref{fig:diverging} shows a diverging
palette in the HSV space with colors ranging from a saturated red $(0, 100, 100)$
over a neutral white $(H, 0, 100)$ to a saturated blue $(240, 100, 100)$.
Although the palette should be balanced with
respect to colorfulness and brightness, the red colors are perceived
to be more more intense and flashy than the corresponding blue
colors.

@@

\setkeys{Gin}{width=0.7\textwidth}
\begin{figure}
\begin{center}
<<highlight,fig=TRUE,echo=FALSE,height=6,width=7,results=hide>>=
mycol <- rep(grey.colors(2)[2:1], 1:2)
mosaic(t(hospital), 
       mar = c(left = 3.5), 
       labeling_args = list(offset_labels = c(left = 0.5), 
         offset_varnames = c(left = 1, top = 0.5), set_labels = 
         list("Visit frequency" = c("Regular","Less than\nmonthly","Never"))),
       split = TRUE, highlighting = 2, gp = gpar(fill = mycol, col = mycol))
@ 
\caption{Spine plot with highlighting for the hospital data.}
\label{fig:highlight}
\end{center}
\end{figure}

@@

A first step in the exploratory analysis of more complex tables is to
get a quick overview on the data. For this, 
all basic plots can be combined in pairwise displays,
arranged in a matrix similar to scatterplots in a pairs
plot. \index{pairs plot}
The diagonal cells contain the variable names, optionally with
univariate statistics, whereas the off-diagonal cells feature plots whose
variables are implicitly specified by the cells' positions in the matrix. %'
In Figure~\ref{fig:pairs}, the diagonal cells show bar plots for the
distributions of the variables, and the off-diagonal cells mosaic plots
for the corresponding pairs of variables. The plots suggest that
admission differs between male and female students, and between the departments,
and that the proportion of male and female students varies across the departments.
In particular, departments A and B have a higher proportion of male students
and a lower rejection rate compared to the other departments.

@@

The next step is to investigate three-way interactions among the variables.
In this example, we have a binary variable of interest (admission)
that is to be ``explained'' by the others. 
A natural way of representing such a three-way table is to use a mosaic
display, 
splitting first in the explanatory variables department and gender and highlighting
the resulting mosaic with respect to the dependent variable admissions. Here, we
use the vertical splits for both explanatory variables, resulting in the 
\index{doubledecker plot} doubledecker plot 
in Figure~\ref{fig:doubleucb}. From the widths of the
tiles it can be seen that students apply to the six departments in
unequal numbers (with a particularly small
number of females in departments~A and B),
and from the highlighting we see again that the admission rates differ
among the departments (roughly speaking, the admission rate is high for A and B,
low for F and in between for C to E). The rates 
are equal for male and female students, except for department A where
\textit{more} female than male students are admitted.

@@

In a first step, we create a mosaic plot for an exploratory view of the data:
Attitude towards punishment clearly is the dependent variable 
here and should be used as the last split via highlighting. 
Furthermore, we expect age and education to have an influence on both memory
and attitude, hence they should be used first for splitting as stratifying
variables. In fact, the question is whether memory has some influence on attitude, given 
age and education. We choose to create a three-way mosaic plot of the 
hypothesized explanatory variables (first splitting horizontically by age, 
then vertically by education, and finally horizontically again by memory), 
and to have attitude, the dependent variable, 
highlighted in the tiles (see Figure~\ref{fig:punishhigh}).
We can see that half of the people are more than 40 years of
age, most of whom of only completed elementary school (much more than 
in the other age groups). Furthermore, it can be seen that the proportion
of people with memories of punishments increases with age and that
the approval rate decreases with education (this, however, is better shown in 
Figure~\ref{fig:punishcotab} described below). As for the question whether
attitude depends on memory, the plot suggests quite clearly (and somewhat
surprisingly) that for those people with elementary school education,
a higher proportion \emph{with} memories tends to
accept moderate punishment of children than those without: 
experience of punishment seems to support use of punishment.
For the other education groups, the picture is less clear: some cells 
indicate the same association, while others do not.

@@

In summary, both age and education clearly matter here.
But is there really any significant influence of memory on attitude towards 
punishment apart from that? Or, rephrased as a hypothesis:
Are memory and attitude conditionally independent, given age and education?
To focus on this question, we take another model-based view of the data
which avoids the distracting influence of the marginal variables (age
and education): We employ
partial mosaic plots in a trellis layout (conditioning on age and education) with
residual-based shading and data-driven cut-offs (see Figure~\ref{fig:punishcotab}).
This visualization brings out much more clearly that the increased approval rate
among people with memories of punishment is present mainly in the elementary education
column and, to a lesser degree, in the age 40+ row. Using a 
permutation test similar to the one discussed in Section \ref{sec:test}, 
model statistics reveal that the hypothesis is rejected at a significance level of 
1\%, but the shading clearly shows that this association is
only significant in two cells (the two older age groups with elementary education)
where fewer people without memories do not approve punishment than expected under
(conditional) independence. In the group of people with age between
25--39, the lighter shading indicates that here
the association is only significant at the 10\% level.

@@

The advantage of the exploratory view 
is to visualize the joint distribution of all variables.
On the other hand, the visualization might be 
strongly influenced by the marginal distribution of education over age (in particular
the large proportion of 40+ people with elementary education),
which is not of interest in the conditional independence problem. The
partial mosaic plot suppresses this effect by complementing the 
conditioning in the model with conditioning in the visualization.

@@

Mosaic, association, and sieve plots can be used for visualizing multi-way tables
by applying them to flat representations. Mosaic plots are particularly useful
for exploratory analysis, whereas the other two require specification of a certain
independence model from which deviations should be examined. In addition, specialized
`flavours' of the plots can leverage
exploratory analyses (pairs plots, highlighting, doubledecker plots) 
or model-based analyses (residual-based shading, conditional plots).
